### 2022-10-05 AISTATS2023/paper.tex
updated method description

\begin{abstract}
Accumulated Local Effects (ALE) is a popular approach to explainable AI that describes how a feature influences the decisions of a model on average, taking into account potential correlations between the features.  However, ALE does not provide an explicit component for modeling the uncertainty of the effect, i.e., the level of heterogeneity of the effects behind the average explanation, which is crucial for a complete interpretation when the black-box function contains interactions between the correlated features.  In this work, we propose Uncertainty-aware ALE (UALE), an extension of ALE for quantifying the heterogeneous effects. We show the importance of bin-splitting for a robust approximation of the uncertainty and we formally prove the conditions for an unbiased estimation. We finally propose a computationally-grounded algorithm for finding the optimal sequence of bins given the limited instances of the training set. We demonstrate through synthetic and real datasets, that UALE quantifies the average and heterogeneous effects correctly and approximates them robustly by optimizing the bin-splitting procedure.
\end{abstract}

@@

In this paper, we present Uncertainty-aware Accumulated Local Effects
(UALE) an extension of ALE for measuring the uncertainty of the
explanation, i.e.~how certain we are that the average explanation would predict the feature effect to an instance drawn at random. We show that choosing
an appropriate sequence of intervals is particularly important for a
robust estimation of the uncertainty, providing an algorithm that
automates this step. We also provide theoretical proofs for an
unbiased estimation of the uncertainty. We apply our method in multiple
synthetic and real world datasets for evaluating its performance.

@@

We extend ALE definition of Eq.~(\ref{eq:ALE}) with a component for
quantifying the uncertainty. We denote as \(\mathcal{H}(z)\) the
uncertainty of the local effects at a specific point \(x_s=z\) and we
quantify it as the standard deviation of the local explanations:
\begin{equation}
  \label{eq:ALE_var}
  \mathcal{H}(z) := \sigma(z) = \sqrt{\mathbb{E}_{\Xcb|z}\left [ \left (\dfdx (z, X_c) - \mu(z) \right )^2 \right ] }
\end{equation}
\noindent
The uncertainty emerges from the natural characteristics of the
experiment, i.e.,~the feature correlations existent in the data
generating distribution and the implicit interactions of the black-box
function. We also define the accumulated uncertainty at \(x_s\), as
the accumulation of the standard deviation of the local effects along
the axis:
\begin{equation}
  \label{eq:ALE_acc_unc}
  f^{\mathtt{ALE}}_{\sigma}(x_s) = \int_{x_{s, min}}^{x_s} \sigma(z) \partial z
\end{equation}
\noindent
UALE formulates the effect at a specific point \(x_s\) with a compact
tuple that consists of the average effect and the uncertainty
\((\mu(z), \sigma(z))\) and visualizes them as a continuous curve with
a confidence region, i.e.
\(f^{\mathtt{UALE}}(x_s) := f^{\mathtt{ALE}}_{\mu}(x_s) \pm
f^{\mathtt{ALE}}_{\sigma}(x_s)\).

@@

In real scenarios, all estimations are based on the instances of the
training set. It is impossible to estimate \(\mu(z), \sigma(z)\) at
the granularity of a point, because the probability of observing a
sample inside the region \([x_s - h, x_s + h]\) tends to zero, when
\(h \to 0\). Therefore we define the regional-effect \(\mu(z_1, z_2)\)
and the regional-uncertainty \(\sigma(z_1, z_2)\), as the summary
statistics that characterize the effect and the uncertainty inside the
interval \([z_1, z_2)\):

\begin{equation}
  \label{eq:mu_bin}
  \mu(z_1, z_2) = \frac{1}{z_2 - z_1} \int_{z_1}^{z_2}
  \mathbb{E}_{\xc|z}\left [f^s(z, \xc) \right ] \partial z
\end{equation}

\begin{equation}
  \label{eq:var_bin}
  \sigma(z_1, z_2) = \sqrt{\frac{\int_{z_1}^{z_2}
  \mathbb{E}_{\xc|x_s=z} \left [ (f^s(z, \xc) - \mu(z) )^2 \right] \partial z}{z_2 - z_1}}
\end{equation}

%
Intuitively, the regional-effect and the regional-uncertainty are the
average effect \(\mu(z^*)\) and the uncertainty \(\sigma(z^*)\) that I
expect to get if draw a sample at random from the region
\([z_1, z_2]\) given a uniform distribution
\(z^* \sim \mathcal{U}(z_1, z_2)\).

If we also define as \(\mathcal{Z}_s\) the sequence of \(K+1\) points
that partition the \(s\)-th feature axis into \(K\) variable-size
intervals, i.e. \(\mathcal{Z}_s = \{z_0, \ldots, z_K\}\), we can
redefine UALE using an interval-based formulation
\(\tilde{f}^{\mathtt{UALE}}(x_s):= \tilde{f}^{\mathtt{ALE}}_{\mu}(x_s)
\pm \tilde{f}^{\mathtt{ALE}}_{\sigma}(x_s)\), where:

\begin{equation}
  \label{eq:ALE_2}
  \tilde{f}^{\mathtt{ALE}}_{\mu}(x_s) = \sum_{k=1}^{k_x} \mu(z_{k-1}, z_k) (z_k - z_{k-1})
\end{equation}

\begin{equation}
  \label{eq:ALE_accumulated_var}
  \tilde{f}^{\mathtt{ALE}}_{\sigma}(x_s) =  \sum_{k=1}^{k_x} \sigma(z_{k-1}, z_k) (z_k - z_{k-1})
\end{equation}
%

@@

Eq.\eqref{eq:2} is an unbiased estimator of Eq.~(\ref{eq:mu_bin})
under the assumption that the points are uniformly distributed inside
the intervals. The same does not hold in the general case for
Eq.\eqref{eq:3}. In the general case, \(\hat{\sigma}^2(z_{k-1}, z_k)\)
is an unbiased estimator of the \textit{observable} standard deviation
\(\sigma_{obs}(z_1, z_2) = \sqrt{\frac{\int_{z_1}^{z_2}
    \mathbb{E}_{\xc|x_s=z} \left [ (f^s(z, \xc) - \mu(z_1, z_2) )^2
    \right] \partial z}{z_2 - z_1}}\), which as we prove in Theorem 1
is an overestimation of \(\sigma(z_1, z_2)\).

@@

\begin{equation}
    \label{eq:bin-uncertainty-proof}
 \sigma_{obs}^2(z_1, z_2) = \sigma^2(z) + \frac{\int_{z_1}^{z_2}\rho^2(z) \partial z}{z_2 - z_1}
\end{equation}
\noindent
Theorem 1 reveals an important connection between the ground-truth and
the observable uncertainty inside a region. Setting as ground-truth
the regional uncertainty
\(\mathcal{H}_{true}(z_1, z_2) = \sigma(z_1, z_2)\) and as observable
uncertainty the one we estimate with Eq.~(\ref{eq:3}), i.e.
\(\mathcal{H}_{obs}(z_1, z_2) := \sigma_{obs}(z_1, z_2)\), then there
is an error term quantified by the residual inside the interval, i.e.
\(\mathcal{E}(z_1, z_2) = \sqrt{\frac{\int_{z_1}^{z_2}\rho^2(z)
    \partial z}{z_2 - z_1}}\), such that:

\begin{equation}
  \label{eq:uncertainty-bin}
  \mathcal{H}^2_{obs}(z_1, z_2) = \mathcal{H}^2_{true}(z_1, z_2) + \mathcal{E}^2(z_1, z_2)
\end{equation}
%
Eq.~(\ref{eq:uncertainty-bin}) shows that the error term quantifies
how much the observable uncertainty of the region \([z_1, z_2]\)
deviates for the one we would like to measure and provides a
quantified metric to evaluate a particular splitting. As we prove in
the Appendix, UALE uncertainty approximation is unbiased if and only if the
partitioning is such that ther error term is zero for all bins, i.e. 
\(\sum_{k=1}^K \mathcal{E}^2(z_{k-1}, z_{k}) = 0\)

@@

In this section we formulate bin-splitting as an unsupervised
clustering problem. The main idea of the clustering problem is to find
the best bin-splitting solution, handling appropriately two
conflicting objectives. On the one hand, we want to minimize the error
term \(\mathcal{E}\) defined above for minimizing the error in the
uncertainty estimation and on the other hand, we want wide-enough bins
for including a fair population of samples for a robust estimation of
\(\hat{\mu}(z_1, z_2), \hat{\sigma}(z_1, z_2)\). We set-up the
following optimization problem:

\begin{equation}
  \label{eq:opt}
\begin{aligned}
  \min_{ \mathcal{Z} = \{z_0, \ldots, z_K\}} \quad & \mathcal{L} = \sum_{k=1}^K \tau_k \hat{\sigma}^2(z_{k-1}, z_k) \Delta z_k \\
  \textrm{where} \quad & \Delta z_k = z_k - z_{k-1} \\
  & \tau_k = 1 - \alpha \frac{|S_k|}{N} \\
  \textrm{s.t.} \quad & |\mathcal{S}_k| \geq N_{\mathtt{NPB}}\\
                                     & z_0 = x_{s,min}\\
                                     & z_K = x_{s, max}
\end{aligned}
\end{equation}
%
We search for the sequence of intervals \(z_0, \ldots, z_K\) that
minimizes the sum of bin costs. The cost of each bin is the
approximated variance \(\hat{\sigma}^2_k\) scaled by the bin length
\(\Delta z_k\) and discounted by the term \(\tau_k\). The term
\(\tau_K\) acts as a tie-braker that favors the selection of a bigger
bin in case it has similar variance with the aggregate variance of
many smaller ones. The constraint of at least \(N_{\mathtt{PPB}}\)
points per bin sets the lowest-limit for a \textit{robust}
estimation. The user can choose to what extent they favor the creation
of wide bins through the discount parameter by setting the parameter
\(\alpha\) and where they set the minimum for robust approximation
with the parameter \(N_{\mathtt{PPB}}\). For providing a rough idea of
some sensible value, in our experiments we set \(\alpha = 0.2\) which
means that the discount ranges in \([0\%, 20\%]\) and
\(N_{\mathtt{PPB}} = \frac{N}{20}\). It is also important to clarify
that by minimizing \(\hat{\sigma}^2_k \approx \mathcal{H}_{obs}^2\),
we actually minimize the squared error term \(\mathcal{E}^2\), since
the term \(\mathcal{H}^2_{true}\) is independent of the bin splitting
procedure.

>>>>

\begin{abstract}
  Accumulated Local Effects (ALE) is a popular approach to explainable
  AI that describes how a feature influences the decisions of a model
  on average, taking into account potential correlations between the
  features.  However, ALE does not provide an explicit component for
  modeling the uncertainty of the effect, i.e., the level of
  heterogeneity of the effects behind the average explanation, which
  is crucial for a complete interpretation when the black-box function
  contains interactions between the correlated features. In this work,
  we define Uncertainty-aware ALE (UALE), an extension of ALE for
  quantifying the heterogeneous effects and we propose a method for
  estimating it. We formally prove the conditions the bin-splitting,
  i.e.~the partitioning the domain of the feature of interest into
  consecutive non-overlapping intervals, must respect for an an
  unbiased estimation of the uncertainty. We finally propose a
  computationally-grounded algorithm for finding the optimal sequence
  of bins given the limited instances of the training set. We
  demonstrate through synthetic and real datasets, that UALE
  quantifies the average and heterogeneous effects correctly and
  approximates them robustly by optimizing the bin-splitting
  procedure.
\end{abstract}

@@

In this paper, we present Uncertainty-aware Accumulated Local Effects
(UALE) an extension of ALE for measuring the uncertainty of the
explanation, i.e.~how certain we are that the average explanation
would predict the feature effect to an instance drawn at random. We
show that choosing an appropriate sequence of intervals is
particularly important for a robust and unbiased estimation of the
uncertainty and we provide an algorithm that automates this step. We
also formally prove the bin-splitting conditions for an unbiased
estimation of the uncertainty. We apply our method in multiple
synthetic and real world datasets for evaluating its performance.

@@

We extend ALE definition of Eq.~(\ref{eq:ALE}) with a component for
quantifying the uncertainty. We denote as \(\mathcal{H}(z)\) the
uncertainty of the local effects at a specific point \(x_s=z\) and we
quantify it as the standard deviation of the local explanations
\(\mathcal{H}(z) = \sigma(z)\), where:

\begin{equation}
  \label{eq:ALE_var}
  \sigma^2(z) = \mathbb{E}_{\Xcb|z}\left [ \left (\dfdx (z, X_c) - \mu(z) \right )^2 \right ] 
\end{equation}
\noindent
The uncertainty emerges from the natural characteristics of the
experiment, i.e.,~the feature correlations existent in the data
generating distribution and the implicit interactions of the black-box
function. We also define the accumulated uncertainty at \(x_s\), as
the accumulation of the standard deviation of the local effects along
the axis:
\begin{equation}
  \label{eq:ALE_acc_unc}
  f^{\mathtt{ALE}}_{\sigma}(x_s) = \int_{x_{s, min}}^{x_s} \sigma(z) \partial z
\end{equation}
\noindent
UALE formulates the effect at a specific point \(x_s\) with a compact
tuple that consists of the average effect and the uncertainty
\((\mu(z), \sigma(z))\) and visualizes them as a continuous curve with
a confidence region, i.e.
\(f^{\mathtt{UALE}}(x_s) := f^{\mathtt{ALE}}_{\mu}(x_s) \pm
f^{\mathtt{ALE}}_{\sigma}(x_s)\).

@@

In real scenarios, all estimations are based on the instances of the
training set. Therefore, we cannot estimate \(\mu(z), \sigma(z)\) at
the granularity of a point, because the probability of observing a
sample inside the region \([x_s - h, x_s + h]\) tends to zero, when
\(h \to 0\). We define the regional-effect \(\mu(z_1, z_2)\) and the
regional-uncertainty \(\mathcal{H}(z_1, z_2) = \sigma(z_1, z_2)\), as the summary statistics
that characterize the effect and the uncertainty inside the interval
\([z_1, z_2)\):

\begin{equation}
  \label{eq:mu_bin}
    \mu(z_1, z_2) = \mathbb{E}_{z \sim \mathcal{U}(z_1,z_2)} [\mu(z)]
    = \frac{\int_{z_1}^{z_2} \mu(z) \partial z}{z_2 - z_1}
\end{equation}

\begin{equation}
  \label{eq:var_bin}
  \sigma^2(z_1, z_2) = \mathbb{E}_{z \sim \mathcal{U}(z_1,z_2)} [\sigma^2(z)] =  \frac{\int_{z_1}^{z_2} \sigma^2(z)  \partial z}{z_2 - z_1}
\end{equation}

%
Intuitively, the regional-effect and the regional-uncertainty quantify
the expected average effect and the expected uncertainty if we
randomly draw a point \(z^*\) given a uniform distribution
\(z^* \sim \mathcal{U}(z_1, z_2)\). If we also define as
\(\mathcal{Z}\) the sequence of \(K+1\) points that partition the
domain of the \(s\)-th feature into \(K\) variable-size intervals,
i.e.  \(\mathcal{Z} = \{z_0, \ldots, z_K\}\), we can redefine UALE
using an interval-based formulation
\(\tilde{f}^{\mathtt{UALE}}_{\mathcal{Z}}(x_s):= \tilde{f}^{\mathtt{ALE}}_{\mathcal{Z}, \mu}(x_s)
\pm \tilde{f}^{\mathtt{ALE}}_{\mathcal{Z}, \sigma}(x_s)\), where:

\begin{equation}
  \label{eq:ALE_2}
  \tilde{f}^{\mathtt{ALE}}_{\mathcal{Z}, \mu}(x_s) = \sum_{k=1}^{k_x} \mu(z_{k-1}, z_k) (z_k - z_{k-1})
\end{equation}

\begin{equation}
  \label{eq:ALE_accumulated_var}
  \tilde{f}^{\mathtt{ALE}}_{\mathcal{Z}, \sigma}(x_s) =  \sum_{k=1}^{k_x} \sigma(z_{k-1}, z_k) (z_k - z_{k-1})
\end{equation}
%

@@

Eq.\eqref{eq:2} is an unbiased estimator of Eq.~(\ref{eq:mu_bin})
under the assumption that the points are uniformly distributed inside
the intervals. The same does not hold in the general case for
Eq.\eqref{eq:3}. In the general case, \(\hat{\sigma}^2(z_{k-1}, z_k)\)
is an unbiased estimator of the \textit{observable} variance
\(\sigma^2_{obs}(z_1, z_2) = \frac{\int_{z_1}^{z_2}
    \mathbb{E}_{\xc|x_s=z} \left [ (f^s(z, \xc) - \mu(z_1, z_2) )^2
    \right] \partial z}{z_2 - z_1}\), which as we prove in Theorem 1
is an overestimation of \(\sigma^2(z_1, z_2)\).

@@

\begin{equation}
    \label{eq:bin-uncertainty-proof}
 \sigma_{obs}^2(z_1, z_2) = \sigma^2(z_1, z_2) + \frac{\int_{z_1}^{z_2}\rho^2(z) \partial z}{z_2 - z_1}
\end{equation}
\noindent
Theorem 1 reveals an important connection between the ground-truth and
the observable uncertainty inside a region. For convenience, we denote
as
\(\mathcal{E}^2(z_1, z_2) = \frac{\int_{z_1}^{z_2}\rho^2(z) \partial
  z}{z_2 - z_1}\) the error term that models the expected square
residual inside the interval and as
\(\mathcal{H}_{obs}(z_1, z_2) := \sigma_{obs}(z_1, z_2)\) the
observable uncertainty. It holds that the observable uncertainty is an
overestimation of the correct uncertainty

\begin{equation}
  \label{eq:uncertainty-bin}
  \mathcal{H}^2_{obs}(z_1, z_2) = \mathcal{H}^2(z_1, z_2) + \mathcal{E}^2(z_1, z_2)
\end{equation}
%
and, therefore, the estimation is unbiased only in case
\(\mathcal{E}^2(z_1, z_2) = 0\).

@@

In this section we formulate bin-splitting as an unsupervised
clustering problem, searching for a solution that compromises two
contradictory objectives. On the one hand, we want to minimize the
error term \(\mathcal{E}\) for approaching an unbiased estimation and
on the other hand, we want wide-enough bins for including a fair
population of samples for a robust estimation of
\(\hat{\mu}(z_1, z_2), \hat{\sigma}(z_1, z_2)\). We set-up the
following optimization problem:

\begin{equation}
  \label{eq:opt}
\begin{aligned}
  \min_{ \mathcal{Z} = \{z_0, \ldots, z_K\}} \quad & \mathcal{L} = \sum_{k=1}^K \tau_k \hat{\sigma}^2(z_{k-1}, z_k) \Delta z_k \\
  \textrm{where} \quad & \Delta z_k = z_k - z_{k-1} \\
  & \tau_k = 1 - \alpha \frac{|S_k|}{N} \\
  \textrm{s.t.} \quad & |\mathcal{S}_k| \geq N_{\mathtt{NPB}}\\
                                     & z_0 = x_{s,min}\\
                                     & z_K = x_{s, max}
\end{aligned}
\end{equation}
%
We search for the sequence of intervals \(z_0, \ldots, z_K\) that
minimizes the sum of the bin costs. The cost of each bin is the
approximated variance \(\hat{\sigma}^2_k\) scaled by the bin length
\(\Delta z_k\) and discounted by the term \(\tau_k\). The term
\(\tau_K\) favors the selection of a bigger bin in case it has similar
variance with the aggregate variance of many smaller ones. The
constraint of at least \(N_{\mathtt{PPB}}\) points per bin sets the
lowest-limit for a \textit{robust} estimation. The user can choose to
what extent they favor the creation of wide bins through the discount
parameter by setting the parameter \(\alpha\) and where they set the
minimum for robust approximation with the parameter
\(N_{\mathtt{PPB}}\). For providing a rough idea, in our experiments
we set \(\alpha = 0.2\) which means that the discount ranges between
\([0\%, 20\%]\) and \(N_{\mathtt{PPB}} = \frac{N}{20}\). It is also
important to clarify that by minimizing
\(\hat{\sigma}^2_k \approx \mathcal{H}_{obs}^2\), we actually minimize
the squared error term \(\mathcal{E}^2\), since the term
\(\mathcal{H}^2_{true}\) is independent of the bin splitting.
